Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-08-22 TACO-paper-v2/coffee.tex
Minor fixes

%Computational cost is a critical limitation in scientific computing,
%especially for finite element simulations. To provide one particular
%example we are particularly concerned about, it has been well
%established that mesh resolution Consequently, our aggressive
%optimization of local assembly, which may even take up to 60% of the
%overall FEM's execution time, directly impacts the performance of
%large-scale scientific simulations running on supercomputers.

@@

%JUNK: , with synchronization handled at the level of the run-time
%support Computational cost is a critical limitation in finite element
%simulations. High resolution in time and space is critical to the
%level of error in the simulations, but computational cost rises
%superlinearly with resolution. The result of this is that simulations
%are conducted at the resolution that the user can afford (wall time,
%allocated supercomputing units or the direct financial cost of cloud
%time)

% FIXME: citation for Met Office?

The time required to apply the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells to obtain a satisfactory approximation
of the PDE, possibly of the order of trillions, as in
\cite{Rossinelli2013}. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centers have a strict time
limit in which to produce a forecast - 60 minutes in the case of the
UK Met Office. Producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper. We address, in
particular, the well-known problem of optimizing the local assembly
phase of the finite element method
\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
can be responsible for a significant fraction of the overall computation
run-time, often in the range 30-60$\%$. With respect to these studies,
we propose a novel set of composable code transformations targeting, for
the first time, cross-loop arithmetic intensity, with emphasis on instruction-level 
parallelism, redundant computation, and register locality. How such transformations generalize
to domains other than local assembly will also be discussed.

@@

Our experiments show that the original generated code for non-trivial
assembly kernels, despite following state-of-the-art techniques,
% FIXME: what are these state-of-the-art techniques?
remains suboptimal in the context of modern multicore
architectures. Our domain-aware cost-model-driven sequence of code
transformations, aimed at improving SIMD vectorization and register
data locality, can result in performance improvements up to
4.4$\times$ over original kernels, and around 70$\%$ of the test cases
obtain a speed up greater than 2$\times$. The contributions of this
paper are
\begin{enumerate}
\item An optimization strategy for finite element local assembly that
  exploits domain knowledge and goes beyond the limits of both vendor
  and research compilers.
\item The design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item A systematic analysis using a suite of examples of real-world
  importance that is evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \item A full-application evaluation to demonstrate that the optimization
 strategy greatly impacts the whole finite element method.
\end{enumerate}
Our contributions also include a theoretical study involving
\begin{enumerate}
 \item An analysis of the generality of the proposed code transformations.
 \item An investigation of their applicability to different computational
 domains.  
\end{enumerate}
The paper is organized as follows. In Section~\ref{sec:background} we
provide some background on local assembly, show code generated by
Firedrake and emphasize the critical computational
aspects. Section~\ref{sec:code-transf} describes the various code
transformations, highlighting when and how domain knowledge has been
exploited. The design and implementation of our compiler is discussed
in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
shows performance results. Related work is described in
Section~\ref{sec:related-work}. The generality of our optimization strategy
and the applicability to other domains is discussed in Section~\ref{sec:generality}. 
Finally, Section~\ref{sec:conclusions} reviews our contributions in the light 
of our results, and identifies priorities for future work.

>>>>

@@

The time required to apply the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells to obtain a satisfactory approximation
of the PDE, possibly of the order of trillions, as in
\cite{Rossinelli2013}. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centers have a strict time
limit in which to produce a forecast - 60 minutes in the case of the
UK Met Office. Producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper.

We address, in particular, the well-known problem of optimizing the local assembly
phase of the finite element method \cite{Francis,quadrature1, petsc-integration-gpu, Kirby-FEM-opt}, 
which can be responsible for a significant fraction of the overall computation
run-time, often in the range 30-60$\%$. With respect to these studies,
we propose a novel set of composable code transformations targeting, for
the first time, cross-loop arithmetic intensity, with emphasis on instruction-level 
parallelism, redundant computation, and register locality. How such transformations 
generalize to domains other than finite element local assembly will also be discussed.

@@

Our experiments show that the original generated code for non-trivial
assembly kernels, despite following state-of-the-art techniques,
% FIXME: what are these state-of-the-art techniques?
remains suboptimal in the context of modern multicore
architectures. Our domain-aware cost-model-driven sequence of code
transformations, aimed at improving SIMD vectorization and register
data locality, can result in performance improvements up to
4.4$\times$ over original kernels, and around 70$\%$ of the test cases
obtain a speed up greater than 2$\times$. The contributions of this
paper are
\begin{enumerate}
\item An optimization strategy for finite element local assembly that
  exploits domain knowledge and goes beyond the limits of both vendor
  and research compilers.
\item The design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item A systematic analysis using a suite of examples of real-world
  importance that is evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \item A full-application evaluation to demonstrate that the optimization
 strategy greatly impacts the whole finite element method.
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}
Our contributions also include a theoretical study involving
\begin{enumerate}
 \setcounter{enumi}{\theenumTemp}
 \item An analysis of the generality of the proposed code transformations
 and an investigation of their applicability to different computational
 domains.  
\end{enumerate}
The paper is organized as follows. In Section~\ref{sec:background} we
provide some background on local assembly, show code generated by
Firedrake and emphasize the critical computational
aspects. Section~\ref{sec:code-transf} describes the various code
transformations, highlighting when and how domain knowledge has been
exploited. The design and implementation of our compiler is discussed
in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
shows performance results. The generality of our optimization strategy
and the applicability to other domains is discussed in Section~\ref{sec:generality}. 
Related work is described in Section~\ref{sec:related-work}. Finally, 
Section~\ref{sec:conclusions} reviews our contributions in the light 
of our results, and identifies priorities for future work.
